Open archaeology, open source?

### 2023-10-11 analysis/_05-collaboration.qmd
Add some introductory text to section 5

<!-- Results II: description of observed collaborative practices

- How active are projects?
- How many people are involved?
- Conversely, how many projects are individual people involved in?
- What is the level of uptake of GitHub's collaboration features?
- Has this changed over time?

Figures:

- Charts presenting commit rates and repo longevity
- Charts presenting the number of contributors and the amount of work that they each contributed
- Charts presenting the use of issues and comments

-->

But how is this activity distributed across repositories? In this section we assess projects' pulse, or the amount of activity pertaining to a repository.

@@

```{r plot-lifespan}
oarch %>%
  unnest(gh_commits) %>%
  drop_na(datetime) %>%
  group_by(item_name) %>%
  add_tally() %>%
  filter(n > 1) %>%
  mutate(first_commit = min(datetime),
         mid_commit = median(datetime),
         last_commit = max(datetime)) %>%
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  geom_linerange(size = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(limits = c(ymd("2005-01-01", tz = "UTC"),
                              ymd("2021-05-31", tz = "UTC")),
                   date_breaks = "2 years", date_labels = "%Y") +
  # coord_flip() +
  labs(
    title = "Project lifespan",
    subtitle = glue("Individual commits in GitHub repositories (with more than 1 commit)"),
    x = NULL, y = NULL,
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

```{r lifespan-stats}
oarch %>% 
  unnest(gh_commits) %>% 
  drop_na(datetime) %>% 
  group_by(item_name) %>%
  mutate(n_contribs = map_int(gh_contribs, vctrs::vec_size)) %>% 
  summarise(
    category = first(category),
    n_contribs = first(n_contribs),
    n = n(),
    first_commit = min(datetime),
    last_commit = max(datetime),
    lifespan = as.integer(last_commit - first_commit),
    commit_rate = n / lifespan
  ) ->
  oarch_lifespan
```

@@

```{r plot-lifespan-hist}
# TODO: combine with patchwork
# TODO: where does oarch_lifespan come from?
oarch_lifespan %>%
  ggplot(aes(x = lifespan)) +
  geom_histogram(binwidth = 100) +
  scale_x_continuous(breaks = scales::breaks_width(1000)) +
  labs(
    title = "Project lifespan",
    subtitle = "Time between first and latest commit, by repository",
    x = "Days", y = NULL,
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

```{r plot-lifespan-vs-rate}
oarch_lifespan %>%
  ggplot(aes(x = lifespan, y = commit_rate, size = n)) +
  geom_point(shape = 21, fill = NA) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = "Project lifespan vs. average commit rate",
    x = "Lifespan (days)", y = "Commits per day", size = "Total commits",
    caption = NULL
  ) +
  theme_oarch()

oarch_lifespan %>%
  ggplot(aes(x = lifespan, y = commit_rate, size = n_contribs)) +
  geom_point(shape = 21, fill = NA) +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    title = NULL,
    x = "Lifespan (days)", y = "Commits per day", size = "Contributors",
    caption = "Data: open-archaeo.info and GitHub API"
  ) +
  theme_oarch()
```

@@

```{r plot-issues-hist}
# TODO: combine with comments-hist
oarch %>%
  unnest(gh_issues) %>%
  drop_na(title) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Use of issues",
    caption = "Data: open-archaeo.info and GitHub API",
    x = "Number of issues", y = "Repositories"
  ) +
  theme_oarch()
```

```{r plot-comments-hist}
oarch %>%
  unnest(gh_comments) %>%
  drop_na(issue_num) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(
    title = "Use of comments",
    caption = "Data: open-archaeo.info and GitHub API",
    x = "Number of comments", y = "Repositories"
  ) +
  theme_oarch()
```

>>>>

As well as hosting source code, GitHub and other software forges include systems for facilitating collaboration on code and other projects.
The basic collaborative workflow is inherited from git, which allows multiple users to *commit* (see @tbl-glossary for definitions of this and other git terminology used in this section) code to the repository.
A user with *commit access* to a repository can change any of its contents at will, so this is usually reserved for the project maintainer and known, trusted collaborators.
GitHub extends this model with its *pull request* feature, by which any user can offer to contribute code to a repository to which they don't have commit access.
The maintainer may then choose to *merge* (accept) or *decline* the pull request, facilitating contributions from a wider network of collaborators without the need for permission to be sought in advance.

GitHub also facilitates collaboration on broader project management tasks, primarily through its *issues* feature.^[Apart from issues, GitHub has a very wide range of project management and social media-like features, including wikis, discussion forums and 'kanban' boards. We have not analysed the use of these features here.]
Unless a repository's maintainer specifically configures it otherwise, any user can create an issue attached to another user's repository, or comment on an existing issue.
Issues are typically used to log and track bug reports, feature requests, and other comments and suggestions from the project's user base.
GitHub's pull request feature is also implemented via this system – a pull request is a special type of issue.

Users can also engage with repositories and other users with social media-like features such as *starring* a repository, adding it to a public or private list, or *following* a user.
These actions populate a timeline of through which users can see recent activity and discover new projects related to those they have interacted with in the past.^[This timeline was in fact one of the primary ways we compiled open-archaeo.]
While not constituting a direct form of collaboration, these features can facilitate the formation and maintenance of collaborative networks, in the same way that other social media platforms serve other professional networks.

The prototypical collaborative open source project on GitHub has a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through pull requests, and an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
The data we collected on open-archaeo repositories on GitHub show that uptake of its collaboration features is highly uneven, with very few projects resembling the open source ideal.
<!-- TODO: note that we don't know if maintainers actually aimed for that model, or leave that for the discussion? -->

* How active are projects
  * Lifespan and activity rate vary greatly.
  * Average lifespan, longest lifespan
  * % of projects with only one commit, under one month, under three months, under a year
  * NB. when we say lifespan, we mean *development* lifespan
  * Average commit rate, range of commit rates

* Comparison of activity vs. longevity
  * Highly active projects (>1 commit per day) tend to be very long-lived or very short-lived (not many in the middle)
  * Long-lived projects have a range of activity levels, relatively evenly spread
  * The most "successful" (according to the expectations of open source, i.e. long-lived and active) are also those with large numbers of contributors (up to 150).
  * The central tendency of the distribution is c. [1000,0.10], i.e. projects that are about three years old and with ~3 commits a month; they are solo projects, or small groups.

* How many people are involved?
  *  Conversely, how many projects are individual people involved in?

* Overall, what is the level of uptake of GitHub's collaboration features?
  * Has this changed over time?

@@

```{r fig-lifespan}
#| fig-cap: Lifespan of repositories based on the date of first and last commit. Points indicate intermediate commits. Only repositories with more than one commit are included.
oarch %>%
  unnest(gh_commits) %>%
  drop_na(datetime) %>%
  group_by(item_name) %>%
  add_tally() %>%
  filter(n > 1) %>%
  mutate(
    first_commit = min(datetime, na.rm = TRUE),
    mid_commit = median(datetime, na.rm = TRUE),
    last_commit = max(datetime, na.rm = TRUE)
  ) %>%
  ggplot(aes(x = fct_reorder(item_name, last_commit, .desc = FALSE),
             y = datetime,
             ymin = first_commit,
             ymax = last_commit)) +
  # geom_linerange(size = 0.05) +
  geom_point(size = 0.1) +
  scale_y_datetime(limits = c(ymd("2005-01-01", tz = "UTC"),
                              ymd("2021-05-31", tz = "UTC")),
                   date_breaks = "2 years", date_labels = "%Y") +
  # coord_flip() +
  labs(x = NULL, y = NULL) +
  theme_oarch() +
  theme(
    axis.line.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

```{r generate-oarch-lifespan}
oarch %>% 
  unnest(gh_commits) %>% 
  drop_na(datetime) %>% 
  group_by(item_name) %>%
  mutate(n_contribs = map_int(gh_contribs, vctrs::vec_size)) %>% 
  summarise(
    category = first(category),
    n_contribs = first(n_contribs),
    n = n(),
    first_commit = min(datetime),
    last_commit = max(datetime),
    lifespan = as.integer(last_commit - first_commit),
    commit_rate = n / lifespan
  ) ->
  oarch_lifespan
```

@@

```{r fig-lifespan-hist}
#| fig-cap: Project lifespan 
oarch_lifespan %>%
  ggplot(aes(x = lifespan)) +
  geom_histogram(binwidth = 100) +
  scale_x_continuous(breaks = scales::breaks_width(1000)) +
  labs(
    x = "Lifespan (days)", y = "Repositories",
  ) +
  theme_oarch()
```

```{r fig-lifespan-rate}
#| fig-cap: Lifespan, commit rate and contributors to repositories. Note that both scales are logarithmic.
# TODO: add marginal histograms?
oarch_lifespan %>%
  ggplot(aes(x = lifespan, y = commit_rate, size = n_contribs)) +
  geom_point(shape = 21, colour = "black", fill = alpha("black", 0.3)) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Lifespan (days)", y = "Commits per day", size = "Contributors") +
  theme_oarch()
```

@@

```{r fig-issues-comments}
#| fig-cap: Use of issues and comments
# TODO: include zeroes
# TODO: why are there no twos?
fig_issues_hist <- oarch %>%
  unnest(gh_issues) %>%
  drop_na(title) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(x = "Number of issues", y = "Repositories") +
  theme_oarch()

fig_comments_hist <- oarch %>%
  unnest(gh_comments) %>%
  drop_na(issue_num) %>%
  group_by(repo) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = n)) +
  geom_histogram() +
  scale_x_log10() +
  labs(x = "Number of comments", y = "Repositories") +
  theme_oarch()

fig_issues_hist + fig_comments_hist
```
